Living on the edge: crop boundary effects on precision crop yields in Alberta, Canada

### 2021-10-06 Manuscript/manuscript.tex
Updated figures

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[left=3cm,right=3cm,top=3cm,bottom=3cm]{geometry}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Living on the edge: crop boundary effects on canola and wheat yields in Alberta},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Living on the edge: crop boundary effects on canola and wheat yields in Alberta}
    \author[University of Calgary]{Samuel V. J. Robinson\corref{1}}
   \ead{samuel.robinson@ucalgary.ca} 
    \author[University of Calgary]{Lan H. Nguyen}
   \ead{hoanglan.nguyen@ucalgary.ca} 
    \author[University of Calgary]{Paul Galpern}
   \ead{paul.galpern@ucalgary.ca} 
      \address[University of Calgary]{2500 University Drive NW, Calgary, AB}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Field boundaries can improve crop yields by creating better abiotic conditions for crop growth (e.g.~mosture, temperature), and can act as refuges for beneficial arthropods. This suggests that beneficial crop boundaries may create an intermediate hump-shaped increase in crop yield, where negative edge effects are cancelled out by increased ecosystem services (e.g.~pollination or pest suppression) from the field boundary. However, there is little large-scale evidence showing this, largely because yield is costly and time-consuming to measure. Precision yield data from combine yield monitors has huge potential in this respect, as the equipment is relatively cheap and commonly measured by growers. Average crop yield tended to increase with distance from crop boundaries before plateauing at about 50 m, and yield variation (SD) tended to decrease concurrently with distance. Differences between boundary types tended to be weak and variable, especially for wheat; there was some evidence of certain field boundaries causing a greater decrease in yield at the edge, but these differences tended to be small in magnitude (\textless5\% of yield). Furthermore, there was little evidence of an intermediate increase in yield, suggesting that a) ecosystem services do not strongly depend on distance or b) their effect is relatively small. This study represents one of the first uses of precision yield data to measure ecosystem service provision at large spatial scales.
  \end{abstract}
   \begin{keyword} precision yield; spillover; landscape;\end{keyword}
 \end{frontmatter}

@@

Intensive agricultural production has increased over the last 100 years, and agricultural land now makes up over a third of ice-free land on Earth.
This has allowed increases in human population and increased global stability in production.
However, this is not without cost, as higher-diversity non-crops are converted to lower-diversity crops, resulting in loss of habitat and overall biodiversity of non-target organisms.
Maintaining both biodiversity and production in agroecosystems represents a goal of conservationists and agronomists, and holds the potential for win-win scenarios.

Key to this goal is the preservation of semi-natural land (SNL) in and around crops, which represents the interface between crops and non-crops within agroecosystems.
SNL in agroecosystems is important for both agricultural production and conservation (Prieto-Benítez \& Méndez 2011), and can create better biotic and abiotic conditions for crop growth (Case \emph{et al.} 2019; Weninger \emph{et al.} 2021; but see Lowe \emph{et al.} 2021).
SNL consitutes habitat for mobile organisms, and can act as sources of ecosystem services such as pollination or pest control (Albrecht \emph{et al.} 2020; Gardner \emph{et al.} 2021), but this can vary depending on the system (Gray \& Lewis 2014; Quinn \emph{et al.} 2017).
They also can create microclimate effects that reduce extreme temperature, trap moisture, and reduce wind speed (Kort 1988; Brandle \emph{et al.} 2004).
North American agroecosystems have larger fields, different crop varieties, and agronomic practices, all of which could negate or confound potential effects of SNL.

SNL may affect yields at intermediate distances, depending on the spatial scale at which ecosystem services operate.
Edge effects cause low yields at the edge of crops because of sparse or late seedling emergence, poor microclimate, and competition with weeds.
At the same time, the centre of large fields will not receive ecosystem services if they decay with distance from edges (Kowalchuk \& Jong 1995).
For example, pollination services from central place foragers (bees) that nest in SNL but forage in crops drops rapidly with distance (Garibaldi \emph{et al.} 2011).
Therefore, yield may be maximized at intermediate distances, where the ecosystem services cancel out negative edge effects {[}Van Vooren \emph{et al.} (2017); Figure \ref{fig:hypotheses}{]}.
This also suggests a ``goldilocks'' field size, where negative edge effects are canceled out by ecosystem services.

@@

Studies of SNL effects on crop yield also suffer from limited scope (e.g.~few crop types) and small sample sizes, limiting inference and reducing generality (Baker \emph{et al.} 2018; but see also Redhead \emph{et al.} 2020).
Meta-analyses have been the standard solution to this problem (e.g. Albrecht \emph{et al.} 2020; Lowe \emph{et al.} 2021), but precision yield data holds enormous promise because of its common usage and relatively cheap operating costs.
However, its use is limited for several reasons, namely: 1) Lack of standardized formats between equipment types, 2) Sensor calibration required for field-level accuracy, and 3) low familiarity with spatial statistics.
Ecosystem services can influence both the mean and variability of yield in agroecosystems (Redhead \emph{et al.} 2020); typically only averages (means) are considered, but higher stability (lower variance) in yield can also be valuable from a grower's perspective.
There are few studies of yield variability (only those with large datasets), but precision yield data opens up the possibility of modeling within-field variability, as well as average yield.

In this paper we ask:
1. How does crop yield change with distance from the edge of field?
2. Does this depend on type of field edge?
3. Is there an intermediate distance where yield is maximized or variance is minimized?
To our knowledge, no other study has addressed the effect of field boundaries on productivity using precision yield, making this study highly novel.

@@

Precision yield data were collected directly from farmers across Alberta.
Farmers were solicited for yield data through local agronomists, and we received 298 field-years of data from 5 growers across a total of 7 years (2014-2020).
We converted raw data to a standard csv format using Ag Leader SMS.
85\% of the crop types where either wheat (\emph{Triticum aestivum}), canola (\emph{Brassica napus}), or peas (\emph{Pisum sativum}), three of the most common crops in rotation in Alberta (Statistics Canada 2014).
The remaining crop types had low replication, so we constrained our analysis to only field-years containing wheat (94), canola (119), or peas (39), for a total of 252 field-years of data.
Individual fields contained between 1 and 6 years of data (mean: 3.1), containing a total of 14.4 million data points.

Yield data was collected in rectangles of the same length as the data interval (distance = combine ground speed \(\times\) interval, typically 1 second) and the same width as the combine header (5-7 m, adjusted to account for overlap with adjacent swaths).
We extracted the size of each polygon (m\(^2\)), dry yield (tonnes), and the spatial location, and the sequence of collection (1 - end of harvest)
Because of the large number of yield rectangles per field (30-800 thousand), we used the centroid of each polygon as its location, treating areal data as point data
Seeding and application rates were constant across fields, so we did not consider inputs in our analysis
We used dry yield (tonnes of seed/hectare minus recorded crop moisture) as our measure of crop yield
Precision yield data can be highly variable and is prone to extreme outliers (both positive and negative), especially when combine ground speeds are low and at the beginning of rows
Therefore, we filtered data that lay outside the 95th percentiles of yield within each field, values less than or equal to zero, and values outside of the 95th percentiles of ground speed.
Due to the large number of data at each field, we sub-sampled to 50,000 data points per field to reduce computation time.

@@

Yield data from each field contains both random and systematic errors that must be modeled in order to reveal underlying changes in yield.
We fit an additive model of the effect of boundary distance on crop yield while accounting for within-field spatial variation and temporal variation in the data.
Crop yield can vary within a field due to soil conditions, moisture, seeding rates, herbicide application, and previous agricultural practices.
Ground speed is extremely important to yield monitor accuracy (Arslan \& Colvin 2002), with low ground speed registering higher yields.
Sensor calibration can reduce combine-level bias (such as a combine recording consistently higher/lower yields across fields), this does not address sensor drift that occurs over time.
This may be caused by sensors accumulating debris during harvest (pers. comm. Trent Clark), leading to changes in accuracy and bias over time.
To model this, we fit the following model to each field-year of data:

\begin{equation}
  \begin{split}
  \sqrt{yield} \sim & Normal (\mu, \sigma)\\
  \mu = & Intercept + log(\text{Polygon Size}) + f(\text{Edge Distance}, b=12)_i + \\
   & f(\text{Easting}, \text{Northing}, b=60) + f(\text{Sequence}, b=60) \\
  log(\sigma) =  & Intercept + log(\text{Polygon Size}) + f(\text{Edge Distance}, b=12)_i + \\
   & f(\text{Easting}, \text{Northing}, b=60) + f(\text{Sequence}, b=60) \\
  \end{split}
  \end{equation}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Polygon Size = distance traveled \(\times\) width of header bar (\(m^2\))
\item
  Edge Distance = distance from field edge type \emph{i} (m)
\item
  Easting, Northing = distance from centre of field (m)
\item
  Sequence = order of harvest within field (1--N points)
\item
  f(x,b) = penalized thin-plate regression spline, where \emph{x} is the predictor and \emph{b} is the number of basis dimensions
\end{enumerate}

In addition to modeling edge effects (our variable of interest), this model also accounts for a) differences in harvested area, b) within-field spatial variation not related to edges, and d) shifts in combine accuracy during harvest.
Spatial or temporal variation can be modelled using a Gaussian Process Model (e.g.~Kriging, SPDE Approximations) but this was computationally unfeasible with 50,000 data points per field.
Penalized splines offer a compromise, as they account for nonlinear ``wiggly" relationships in the same way as Gaussian processes but using a reduced number of dimensions and computation time.
The number of basis dimensions was checked with the \emph{gam.check} function from \emph{mgcv}.
The relationship between polygon size and yield was modeled with a log-linear relationship with a single slope term, as this closely matched smoothed versions.
Because polygon size is not intuitive, we converted it into ground speed by regressing speed on distance traveled per polygon, then using this to back-calculate the ground speed for a given polygon size (given a fixed width of header).
All models were fit in \emph{R} using the \emph{mgcv} library (version 1.8.36, Wood 2017), and figures were created with \emph{ggplot2} and \emph{ggpubr} (versions 3.3.3, Wickham 2016; and 0.4.0, Kassambara 2020).

Despite the use of spatial and temporal smoothers, the residuals in most of the models displayed a large amount of spatial and temporal autocorrelation.
\emph{Not really sure how we can get around this. In practice, this means that SEs for all coefficients are too small (i.e.~we gathered 1000 points, but because they're spatially related, it's only equivalent to 100 points).}

To consider results from all field-level models together, we fit models independent of each other and fit an ``overall'' smoother as averages of the field-level smoothers.
However, this does not account for uncertainty in the field-level smoothers, so we used an approach similar to bootstrapping of hierarchical mixed effects models:
1. Extract single posterior sample (\emph{rnorm} in R) of smoother parameters from each field-level model using coefficient estimates and standard error
2. Use posterior sample to create a new smoother from each field
3. Fit new model of new smoothers from all fields, and save this ``meta-smoother''
4. Repeat 1000 times, and calculate coverage intervals (5-95\% percentiles) on saved meta-smoothers
This gives coverage intervals (CIs) for the ``average'' smoother while accounting for field-level variability, and is conceptually similar to a meta-analysis.
Spatial variability was not averaged between fields because differences in field geometry make it impossible to compare fields.

@@

The filtered data from each field still contained a large amount of variation, but the non-stationary additive models were able to reveal underlying yield patterns.
To illustrate this, we show the results from a single wheat field, where we show the relative impact of ground speed, harvest sequence, boundary distance, and spatial variation on both yield mean and yield SD (``patchiness'').
Figure \ref{fig:rawData} shows the raw data collected from a single field, demonstrating a large amount of variation even among the filtered data.
Combine ground speed had a large effect on both yield mean and SD, and there was also a strong effect of harvest sequence on yield mean and SD (Figure \ref{fig:orderSmooths}.
Yield mean and SD tended to decrease with distance from field boundaries, but there were no consistent differences between boundary types (Figure \ref{fig:distSmooths}.
Finally, the random spatial smoothers in Figure \ref{fig:spatialSmooths} show that yield mean and SD systematically vary within fields, likely due to moisture, elevation, or soil conditions (Figure \ref{fig:spatialSmooths}.

\begin{figure}
\includegraphics[width=0.75\linewidth]{../Figures/ExamplePlots/rawData} \caption{Example output from a single field.}\label{fig:rawData}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/ExamplePlots/orderSmooths} \caption{Ground speed and harvest sequence (random temporal variation) smoothers from a single field.}\label{fig:orderSmooths}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/ExamplePlots/distSmooths} \caption{Distance smoothers from a single field.}\label{fig:distSmooths}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/ExamplePlots/spatialSmooths} \caption{Spatial (random spatial variation) smoothers from a single field.}\label{fig:spatialSmooths}
\end{figure}

@@

Across all fields, Standard (grassy border) field boundaries were the most common boundary type, followed by Shelterbelts and Other Crops (Figure \ref{fig:boundaryTypes}).

@@

\hypertarget{canola-fields}{%
\subsubsection{Canola fields}\label{canola-fields}}

Overall, there was a strong negative effect of combine ground speed on both the average and variability of canola yield.
At the field level, both yield average and yield variability (SD of yield) of yield were extremely spatially (average p-values: 2e-16, 1.77e-05) and temporally dependent (2e-16, 0.000243), meaning that average and variability changed with location in the field, as well as sequentially during the harvest.
However, there was no difference in average or variability with harvest sequence when all fields were considered, meaning that combines did not appear to distort yield data from start to finish in any systematic way.

Average canola yield tended to increase with distance from Standard and Other Crop boundaries, leveling off at about 25 m (Figure \ref{fig:canolaPlot} top row, Other crop and Standard panels); yield was approximately 0.1 T/ha (1.5 bu/ac) lower at the edge for both of these boundary types.
Interestingly, Shelterbelts were well-represented in the data (Figure \ref{fig:boundaryTypes}) but the negative edge effect was not strong (Figure \ref{fig:canolaPlot} top row, Shelterbelt panel), indicating that they may provide better micro-climates for crops located near them.
Other boundary types were less common in the data and had correspondingly weaker effects on mean yield, but all showed a small negative edge effect.
Yield variability tended to decrease with distance from the edge of the field, indicating that edge effects contribute to yield ``patchiness'' as well as averages.
At Standard, Bare, and Wetland field boundaries, variability decreased until about 100 m, after which variability was constant, while the pattern was less consistent in Grassland, Other Crop, and Shelterbelt.

\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/ModelSummary3a_canola} \caption{Field boundary effect on canola yield, accounting for the effect of combine speed, spatial variation, and harvest sequence. Upper panel represents mean yield, while the lower panel represents yield variation (i.e. "patchiness").}\label{fig:canolaPlot}
\end{figure}

\hypertarget{wheat-fields}{%
\subsubsection{Wheat fields}\label{wheat-fields}}

Similar to canola, there was a strong negative effect of ground speed on both yield and variability of wheat yield (Figure \ref{fig:wheatPlotSpeed}, left panels).
At the field level, both yield average and yield variability (SD of yield) of yield were spatially (average p-values: 2e-16, 2e-16) and temporally dependent (2e-16, 2e-16), but harvest sequence did not distort the average or SD of wheat yield in a systematic way (Figure \ref{fig:wheatPlotSpeed}, right panel).

Average wheat yield increased slightly with distance from the field boundary, but this was much more variable than the patterns seen in canola yield (Figure \ref{fig:wheatPlot}).
There was also a slight increase in variability away from the field boundary, at least for Standard and Other Crop boundaries, but this was inconsistent among other boundary types.

\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/ModelSummary3a_wheat} \caption{Field boundary effect on wheat yield, accounting for the effect of combine speed, spatial variation, and harvest sequence. Upper panel represents mean yield, while the lower panel represents yield variation (i.e. "patchiness").}\label{fig:wheatPlot}
\end{figure}

\hypertarget{pea-fields}{%
\subsubsection{Pea fields}\label{pea-fields}}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[left=3cm,right=3cm,top=3cm,bottom=3cm]{geometry}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Living on the edge: crop boundary effects on precision crop yields in Alberta, Canada},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Living on the edge: crop boundary effects on precision crop yields in Alberta, Canada}
    \author[University of Calgary]{Samuel V. J. Robinson\corref{1}}
   \ead{samuel.robinson@ucalgary.ca} 
    \author[University of Calgary]{Lan H. Nguyen}
   \ead{hoanglan.nguyen@ucalgary.ca} 
    \author[University of Calgary]{Paul Galpern}
   \ead{paul.galpern@ucalgary.ca} 
      \address[University of Calgary]{2500 University Drive NW, Calgary, AB}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Field boundaries can improve crop yields by creating better abiotic conditions for crop growth (e.g.~moisture, temperature), and can act as refuges for beneficial arthropods. This suggests that beneficial crop boundaries may create an intermediate hump-shaped increase in crop yield, where negative edge effects are cancelled out by increased ecosystem services (e.g.~pollination or pest suppression) from the field boundary. However, there is little large-scale evidence showing this, largely because yield is costly and time-consuming to measure. Precision yield data from combine yield monitors has huge potential in this respect, as the equipment is relatively cheap and commonly measured by growers. In this study, we used 252 field-years of yield monitor data from three common crops -- wheat (\emph{Triticum aestivum}), canola (\emph{Brassica napus}), or peas (\emph{Pisum sativum}) -- recorded across Alberta, Canada, and examined how yield varied with distances from common crop boundary types. Average crop yield tended to increase with distance from crop boundaries before plateauing at about 50 m, and yield variation (SD) tended to decrease with distance. There was evidence of an intermediate increase in yield for wheat away from shelterbelts, and a weak increase in canola, but no other boundary type showed evidence of this. This study represents one of the first uses of precision yield data to measure ecosystem service provision at large spatial scales.
  \end{abstract}
   \begin{keyword} precision yield; spillover; landscape;\end{keyword}
 \end{frontmatter}

@@

Intensive agricultural production has increased over the last 100 years, and agricultural land now makes up over a third of ice-free land on Earth (Ramankutty \emph{et al.} 2018).
This has allowed increases in human population and increased global stability in production (Foley \emph{et al.} 2011).
However, this is not without cost, as higher-diversity non-crops are converted to lower-diversity crops, resulting in loss of habitat and overall biodiversity of non-target organisms.
Maintaining both biodiversity and production in agroecosystems represents a goal of conservationists and agronomists, and holds the potential for win-win scenarios ({``multifunctionality''} \emph{sensu} Mitchell \emph{et al.} 2014; Frei \emph{et al.} 2018).
Key to this goal is the preservation of semi-natural land (SNL) in and around crops, which represents the interface between crops and non-crops within agroecosystems.

SNL in agroecosystems is important for both agricultural production and conservation (Prieto-Benítez \& Méndez 2011), and can create better biotic and abiotic conditions for crop growth (Case \emph{et al.} 2019; Weninger \emph{et al.} 2021; but see Lowe \emph{et al.} 2021).
SNL constitutes habitat for mobile organisms, and can act as sources of ecosystem services such as pollination or pest control (Albrecht \emph{et al.} 2020; Gardner \emph{et al.} 2021), but this can vary depending on the crop system (Gray \& Lewis 2014; Quinn \emph{et al.} 2017) and landscape context (Marja \emph{et al.} 2019; Boetzl \emph{et al.} 2020).
They also can create microclimate effects that reduce extreme temperature, trap moisture, and reduce wind speed and soil erosion (Kort 1988; Brandle \emph{et al.} 2004; Wratten \emph{et al.} 2012).
North American agroecosystems have larger fields, different crop varieties, and agronomic practices, all of which could negate or confound potential effects of SNL.
Semi-natural field boundaries (e.g.~grass and flower strips, tree or shrub shelterbelts, road right-of-ways) are of particular interest, as they are extremely common and require less land than field-sized agricultural set-asides.

Field boundaries can cause both positive or negative effects on crops, depending on the spatial scale at which ecosystem services operate.
Low yields near may occur near the edge of the field because of patchy seed emergence, poor microclimate, shading from trees, or greater competition with weeds.
Field boundaries can also provide helpful ecosystem services, such as pest suppression or pollination, and these also tend to be occur close to the field edge (Duelli \& Obrist 2003).
For example, crop pollination services from bees that nest in field boundaries drops rapidly with distance (Kowalchuk \& Jong 1995; Garibaldi \emph{et al.} 2011).
If negative edge effects of the edge improve with distance and ecosystem services decrease with distance, yield may be maximized at intermediate distances where the ecosystem services cancel out negative edge effects {[}Mitchell \emph{et al.} (2014); Van Vooren \emph{et al.} (2017); Figure \ref{fig:hypotheses}a{]}.
This suggests a ``Goldilocks'' field size, where negative edge effects are canceled out by ecosystem services.
Alternatively, if ecosystem services are constant with distance (or increase, Figure \ref{fig:hypotheses}b) then yield will increase with distance from a field boundary and then plateau.

@@

Boundary effects are difficult to observe without large amounts of crop yield data, which is typically expensive and time-consuming to collect.\\
Studies of semi-natural field boundaries on crop yield tend to suffer from limited spatial scope and small sample sizes, limiting inference and reducing generality (Baker \emph{et al.} 2018; but see also Redhead \emph{et al.} 2020).
Meta-analyses are a common solution to this problem (e.g. Albrecht \emph{et al.} 2020; Lowe \emph{et al.} 2021), but precision yield data holds enormous promise because of its relatively common usage and low operating costs.
However, its use is limited for several reasons, namely: 1) Lack of standardized formats between equipment types, 2) Sensor calibration required for field-level accuracy, and 3) low familiarity with spatial statistics.
Ecosystem services from field boundaries may influence both the mean as well as the variability of yield in agroecosystems (Redhead \emph{et al.} 2020; Hünicken \emph{et al.} 2021); typically only means are considered, but lower variability in yield can also be valuable from a grower's perspective.
There are few studies of yield variability (only those with large datasets), but precision yield data opens up the possibility of modeling within-field variability, as well as average yield.

In this paper use a precision yield dataset from Alberta, Canada to ask the following questions:
1. How does crop yield change with distance from the edge of field?
2. Does this depend on type of field edge?
3. Is there an intermediate distance where yield is maximized or variance is minimized?
To our knowledge, no other study has studied the effects of field boundaries using precision yield data, making this study one of the first of its kind.

@@

Precision yield data were collected directly from farmers across Alberta.
Farmers were solicited for yield data through local agronomists, and we received 298 field-years of data from 5 growers across a total of 7 years (2014-2020).
We converted raw data to a standard csv format using Ag Leader SMS®.
85\% of the crop types where either wheat (\emph{Triticum aestivum}), canola (\emph{Brassica napus}), or peas (\emph{Pisum sativum}), three of the most common crops in rotation in Alberta (Statistics Canada 2014).
The remaining crop types had low replication, so we constrained our analysis to only field-years containing wheat (91), canola (118), or peas (38), for a total of 247 field-years of data.
Individual fields contained between 1 and 6 years of data (mean: 3), containing a total of 13.9 million data points.

\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/Field Locations} \caption{Locations of fields that data were collected from}\label{fig:fieldLocataions}
\end{figure}

Yield data was collected in rectangles of the same length as the data interval (distance = combine ground speed \(\times\) interval, typically 1 second) and the same width as the combine header (5-7 m, adjusted by the combine to account for overlap with adjacent swaths).
We used dry yield (tonnes of seed per hectare minus - crop moisture) as our measure of yield, using the size of each rectangle (m\(^2\)), dry yield (tonnes), and the spatial location.
Because of the large number of yield rectangles per field (30-800 thousand), we used the centroid of each polygon as its location, treating areal data as point data.
Seeding and application rates were constant across fields, so we did not consider inputs in our analysis.
Precision yield data can be highly variable and is prone to extreme outliers, especially when combine ground speeds are low and at the beginning of rows (Griffin \emph{et al.} 2007; Whelan \& Taylor 2013), or when the combine is accelerating or decelerating (Arslan \& Colvin 2002).
Therefore, we used the following filtering procedure to clean the data prior to model fitting:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Removed outliers and ``inliers'' using the spatial filtering procedure from (Vega \emph{et al.} 2019)
\item
  Removed unrealistic yield values (10.75 T/ha for wheat, 8 T/ha for canola and peas), and values of zero or less
\item
  Removed extremely low and high outliers (below 1st percentile, above 99th percentile) for: dry yield, steering angle differences, ground speed
\item
  Removed points with large changes (\textgreater20\%) in ground speed
\item
  Removed extremely low and high changes in position, indicating that the combine had stopped or moved to another part of the field between data collection
\item
  Sub-sampled filtered data to 50,000 data points per field (reduce computation time)
\end{enumerate}

@@

Yield data from each field contains both random and systematic errors that must be modeled in order to reveal underlying changes in yield.
We fit an additive model of the effect of boundary distance on crop yield while accounting for within-field spatial variation.
Crop yield can vary within a field due to soil conditions, moisture, seeding rates, herbicide application, previous agricultural practices, and other factors not directly related to boundary distance.
To model this, we fit the following model to each field-year of data:

\begin{equation}
  \begin{split}
  \sqrt{Yield} \sim & Normal (\mu, \sigma)\\
  \mu = & Intercept + f(\text{Boundary Distance}, b=5)_i + \\
   & f(\text{Easting}, \text{Northing}, b=60)\\
  log(\sigma) =  & Intercept + f(\text{Boundary Distance}, b=5)_i + \\
   & f(\text{Easting}, \text{Northing}, b=60)\\
  \end{split}
  \end{equation}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Yield = weight of crop gathered per area of harvest (T/ha)
\item
  Boundary Distance = distance from field boundary type \emph{i} (m)
\item
  Easting, Northing = distance from centre of field (m)
\item
  Sequence = order of harvest within field (1--N points)
\item
  f(x,b) = penalized thin-plate regression spline, where \emph{x} is the predictor and \emph{b} is the number of basis dimensions
\end{enumerate}

In addition to modeling edge effects (our variable of interest), this model also accounts for within-field spatial variation not related to edges.
Spatiotemporal variation can also be modelled using a Gaussian Process Model (e.g.~Kriging, SPDE Approximations) but this was computationally unfeasible with 50,000 data points per field.
Penalized splines offer a compromise, as they account for nonlinear ``wiggly" relationships in the same way as Gaussian processes but using a reduced number of dimensions and computation time.
We used 5 basis dimensions for the distance smoother and 60 for the spatial smoothers, as we hypothesized that boundary effects would not be as complex as within-field variation (as in Figure \ref{fig:hypotheses}).
All models were fit in \emph{R} using the \emph{mgcv} library (version 1.8.36, Wood 2017), and figures were created with \emph{ggplot2} and \emph{ggpubr} (versions 3.3.3, Wickham 2016; and 0.4.0, Kassambara 2020).

The residuals in most of the models displayed both spatial autocorrelation in the residuals, indicating ``micro-scale'' variation that was not captured by the spatial smoothers (Cressie \& Wikle 2011; Wikle \emph{et al.} 2019).
This is extremely common in precision yield data, and can bias standard errors of regression terms, but because a) there were a large number of data points per field, b) standard errors were relatively small (i.e.~extremely small p-values), and c) results did not change strongly when additional basis functions were added, we decided to leave the models as-is.
Temporal autocorrelation in residuals was also present, but we did not include temporal smoothers because spatial and temporal processes are confounded in precision yield data.
That is, combining of fields occurs in patterns (e.g.~east-to-west) so it is impossible to tell whether changes in yield in certain areas of the field were caused by deviations in the yield monitor sensors over time, or whether they represent changes in growing conditions.
We considered the latter to be a more realistic assumption, so we assigned all random variation as spatial in nature.
Customized combine driving patterns may be able to detect temporal variation, but this is beyond the scope of this research.

To consider results from all field-level models together, we fit models independently and then fit ``overall'' smoothers as averages of the field-level smoothers.
However, this does not account for uncertainty in the field-level smoothers, so we used an approach similar to bootstrapping of hierarchical mixed effects models:
1. Extract single posterior sample (\emph{rnorm} in R) of smoother parameters from each field-level model using coefficient estimates and standard error
2. Use posterior sample to create a new smoother from each field
3. Fit new model of new smoothers from all fields, and save this ``meta-smoother''
4. Repeat 1000 times, and calculate coverage intervals (5-95\% percentiles) on saved meta-smoothers
This gives coverage intervals (CIs) for the ``average'' smoother while accounting for field-level variability, and is conceptually similar to a meta-analysis.
Spatial variability was not averaged between fields because within-field variability is not directly comparable between fields.

@@

The filtered data from each field still contained a large amount of variation, but the non-stationary additive models were able to reveal underlying yield patterns.
To illustrate this, we show the results from a single wheat field, where we show the relative impact of boundary distance and spatial variation on both yield mean and yield SD (``patchiness'').
Figure \ref{fig:spatialSmooths}a shows the raw data collected from a single field, demonstrating a large amount of variation even among the filtered data.
Yield mean and SD tended to decrease with distance from field boundaries, but there were no consistent differences between boundary types (Figure \ref{fig:distSmooths}).
Finally, the random spatial smoothers in Figure \ref{fig:spatialSmooths}b and c show that yield mean and SD systematically vary within fields, likely due to moisture, elevation, or soil conditions.

\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/ExamplePlots/spatialSmooths} \caption{Raw data and spatial smoothers from a single field. Yield averages (means) and variability (log SD) were modeled separately, and both show large spatial dependence within the field.}\label{fig:spatialSmooths}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/ExamplePlots/distSmooths} \caption{Distance smoothers from a single field, showing a negative effect of distance on average yield (first row). The variability smoothers show an intermediate reduction in variability at around 90 m.}\label{fig:distSmooths}
\end{figure}

@@

Across all fields, standard (grassy border) field boundaries were the most common boundary type, followed by shelterbelts and other crops (Figure \ref{fig:boundaryTypes}).

@@

Overall, both yield average and yield variability were extremely spatially dependent across all fields (average p-values both \textless0.0001), meaning that within-field variability was the primary driver of yield across canola, wheat, and pea crops (Table \ref{tab:smoothSummary}, row 6).
Distances from boundary types were also important, but a lower proportion of the fields had strong effects (rows 1-5, cols 3 + 5).

\begin{table}

\caption{\label{tab:smoothSummary}Mean p-values and proportion of p-values less than 0.05 (after Bonneferoni correction) for mean and variability smoothers across all fields.}
\centering
\begin{tabular}[t]{l|r|r|r|r}
\hline
\multicolumn{1}{c|}{} & \multicolumn{2}{c|}{Mean} & \multicolumn{2}{c}{Variability} \\
\cline{2-3} \cline{4-5}
Smoother Type & Average p-value & Proportion <0.05 & Average p-value & Proportion <0.05\\
\hline
Bare & 0.036 & 0.875 & 0.055 & 0.857\\
\hline
Grassland & 0.079 & 0.794 & 0.061 & 0.794\\
\hline
Other Crop & 0.063 & 0.860 & 0.044 & 0.840\\
\hline
Shelterbelt & 0.014 & 0.935 & 0.059 & 0.815\\
\hline
Standard & 0.021 & 0.938 & 0.055 & 0.839\\
\hline
Wetland & 0.045 & 0.833 & 0.013 & 0.857\\
\hline
Spatial Smoother & 0.000 & 1.000 & 0.000 & 1.000\\
\hline
\end{tabular}
\end{table}

Average canola yield tended to increase with distance from all crop boundaries, starting at 2.5 T/ha at the field edge and leveling off at about 3 T/ha (Figure \ref{fig:canolaPlot} top row).
Shelterbelt and standard boundaries were well-represented in the data, and average yield quickly increased with distance away from the field boundary, but shelterbelts had a small intermediate maximum at around 50m before a weak decrease with distance, while average yield away from Standard field boundaries plateaued at 50m and remained roughly constant with distance.
The effect of other boundary types was less consistent, but average yield tended to increase with distance away from grassland, other crop, and wetland, and was roughly the same with distance from bare boundary types.
Yield variability (Figure \ref{fig:canolaPlot} bottom row) was essentially constant with distance for shelterbelt, standard, and other crop field boundaries, but showed some evidence of decreasing with distance from field boundary at wetlands, bare, and grassland.
Interestingly, there was an intermediate reduction in canola yield variability at about 75 m away from Wetland boundaries, indicating a potential stabilizing effect.

\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/ModelSummary3a_canola} \caption{Field boundary effect on canola yield, accounting for the effect of spatial variation. Upper panel represents mean yield, while the lower panel represents yield variation. N refers to number of fields containing this boundary type, and \% refers to the average percentage of field boundary accounted for by this boundary type.}\label{fig:canolaPlot}
\end{figure}
